On the Gravitational stability of gravito-turbulent accretion disks

### 2016-01-14 paper/model3d.tex
regular update

\subsubsection{Beta cooling}
In this case the energy loss per unit volume is specified 
as an explicit function of the thermodynamic variables. 
A prototypical example is 
\begin{align}\label{beta_cool}
  \Lambda =
  %  \frac{P}{(\gamma-1)T}\frac{\left(T-T_\mathrm{irr}\right)}{t_c}, 
  \frac{\mathcal{R}\rho}{(\gamma-1)}\frac{\left(T-T_\mathrm{irr}\right)}{t_c}, 
\end{align}
where 
\begin{align}
  \tcool = \beta\Omega^{-1} 
\end{align}
is the cooling timescale with $\beta$ a constant input parameter, and
\begin{align}
  \tirr = \theta T_\mathrm{eq}
\end{align}
is a reference temperature field parameterized as a constant fraction
$\theta$ of the equilibrium temperature $T_\mathrm{eq}$. We will refer
to $\theta$ as the `irradiation level'.  
%This cooling function restores 
%the disk temperature to $\tirr$ over a time-scale $\tcool$.

   Beta cooling is widely applied in 2D and 3D
numerical simulations of self-gravitating disks \citep{gammie01,
  rice05,baehr15}. It permits numerical experiments to be carried out in a
controlled manner as a function of the cooling time $\beta$.

@@

We remark that for the linear problem  
any explicit cooling function $\Lambda=\Lambda(\rho,T)$ can be
mapped to that obtained from the standard function in Eq.
\ref{beta_cool}. We do this in \S\ref{2dppd} where we adopt a more
realistic beta cooling prescription for 2D PPDs.

\subsubsection{Radiative cooling}\label{rad_cool}
Here we assume energy transfer by radiative diffusion. Then
\begin{align}
  \Lambda &= \nabla\cdot\bm{F},\label{rad_cool1}\\
  \bm{F}   &= -\frac{16\sigma T^3}{3\kappa_d\rho}\nabla T, \label{rad_cool2}
\end{align}
where $\sigma$ is the Stefan-Boltzmann constant and 
$\kappa_d$ is the (dust) opacity. We adopt
\begin{align}\label{opacity_law}
  \kappa_d = \kappa_{d0}T^b,
\end{align}
and take the constant index $b=2$ as appropriate for {\bf blah},
but retain the general notation $b$ to keep track of the opacity.   
%We use radiative cooling when considering 3D PPDs in \S.

In this case we specify a constant viscosity coefficient $\alpha$ and
solve Eq. \ref{vert_eq1}---\ref{thermal_eq}, together with
Eq. \ref{rad_cool1}---\ref{rad_cool2}, as a fourth order system of
ordinary differential equations to obtain $P(z)$, $T(z)$, and hence
$\rho(z)$.

While radiative cooling is arguably more realistic than
parametrized beta cooling, it generally implies a vertically
non-isothermal equilibrium disk, and increases the order of the
linearized equations. It also only applies to optically-thick 
disks. It is possible to modify the flux function to account for
optically-thin disks, but this complication is beyond the scope of
this work.

>>>>

\subsubsection{Beta cooling}
In this case the energy loss per unit volume is specified 
as an explicit function of the thermodynamic variables. 
A prototypical example is 
\begin{align}\label{beta_cool}
  \Lambda =
  %  \frac{P}{(\gamma-1)T}\frac{\left(T-T_\mathrm{irr}\right)}{t_c}, 
  \frac{\mathcal{R}\rho}{(\gamma-1)}\frac{\left(T-T_\mathrm{irr}\right)}{t_c}, 
\end{align}
where 
\begin{align}
  \tcool = \beta\Omega^{-1} 
\end{align}
is the cooling timescale with $\beta$ a constant input parameter, and
\begin{align}
  \tirr = \theta T_\mathrm{eq}
\end{align}
is a reference temperature field, here parameterized as a constant fraction
$\theta$ of the equilibrium temperature $T_\mathrm{eq}$. We will refer
to $\theta$ as the `irradiation level'.  
%This cooling function restores 
%the disk temperature to $\tirr$ over a time-scale $\tcool$.

@@

 Beta cooling is widely applied in 2D and 3D
numerical simulations of self-gravitating disks \citep{gammie01,
  rice05,baehr15}. It permits numerical experiments to be carried out in a
controlled manner as a function of the cooling time $\beta$. 
We remark that for the linear problem  
any explicit cooling function $\Lambda=\Lambda(\rho,T)$ can be
mapped to that obtained from the standard function in Eq.
\ref{beta_cool}. We do this in \S\ref{2dppd} where we adopt a more
realistic beta cooling prescription for 2D PPDs.

\subsubsection{Radiative cooling}\label{rad_cool}
Here we assume energy transfer by radiative diffusion. Then
\begin{align}
  \Lambda &= \nabla\cdot\bm{F},\label{rad_cool1}\\
  \bm{F}   &= -\frac{16\sigma T^3}{3\kappa_d\rho}\nabla T, \label{rad_cool2}
\end{align}
where $\sigma$ is the Stefan-Boltzmann constant and 
$\kappa_d$ is the (dust) opacity. We adopt
\begin{align}\label{opacity_law}
  \kappa_d = \kappa_{d0}T^b,
\end{align}
and take the constant index $b=2$ as appropriate for the cold outer regions
of a realistic PPD \citep{bell94}, but retain the general notation $b$
to keep track of the opacity.    
%We use radiative cooling when considering 3D PPDs in \S.

In this case, we specify a constant viscosity coefficient $\alpha$ and
solve Eq. \ref{vert_eq1}---\ref{thermal_eq}, together with
Eq. \ref{rad_cool1}---\ref{rad_cool2}, as a fourth order system of
ordinary differential equations to obtain equilibrium profiles $P(z)$, $T(z)$, and hence
$\rho(z)$.

While radiative cooling is arguably more realistic than
parametrized beta cooling, it generally implies a vertically
non-isothermal equilibrium disk, and increases the order of the
linearized equations. It formally applies to optically-thick 
disks, but it is possible to modify the flux function to account
for optically-thin disks. However, this complication is beyond the scope of
this work.

### 2016-01-14 paper/paper3d.tex
regular update

\begin{abstract}
Low mass, self-gravitating accretion disks can remain in a  
quasi-steady, `gravito-turbulent' state provided that thermal 
time-scales are sufficiently long, or equivalently for sufficiently
weak turbulent stresses. Otherwise, numerical simulations show that
disk fragmentation ensues. 
We quantify the interpretation of this result as the
compromise of gravitational stability by non-ideal physics: 
\begin{inparaenum}[1)] 
\item cooling, which reduces pressure support; and
\item viscosity, which reduces rotational support. 
\end{inparaenum}
To this end, we analyse the gravitational stability of accretion disks
including these effects explicitly.   
We derive a versatile dispersion relation for linear waves in a 
self-gravitating, viscous, non-adiabatic disk with internal and
external heating, and explicit cooling. We 
consider parameterized cooling functions in two and three dimensions,
as well as full radiative diffusion in 3D. 
We apply linear theory to 
global models of protoplanetary disks (PPDs) subject to gravito-turbulence, modeled as an effective
viscosity, and cooling via dust opacity. We find the sole effect of cooling is insufficient to
destabilize a realistic PPD. However, inclusion of viscosity renders
the disk beyond $\sim 60$AU dynamically unstable on radial 
lengthscales a few times the local disk thickness. This is coincident 
with the numerically-established empirical condition for disk fragmentation based on a
maximum sustainable stress. We 
thus suggest that disk fragmentation can be physically attributed,
at least partially, to the removal of rotational stabilization against
self-gravity by turbulence.   
\end{abstract}
%alternative abstract

\section{Introduction}\label{intro}
Understanding the gravitational stability (GI) of rotating disks   
is fundamental to many astrophysical problems. In the context of 
protoplanetary disks (PPDs), GI has two applications. It can 
provide gravitational torques to transport angular momentum outwards
and thus enable mass accretion inwards \citep{armitage10,turner14}.   
GI may also lead to disk fragmentation, which has been invoked to explain planet formation at
large radii \citep{helled14}. Any realistic study of these phenomena require direct
numerical simulations. Nevertheless, physical insight can be  
obtained through analytical understanding of GI.  %physical insight

@@

\subsection{Viscosity removes rotational support}
It is well known that viscous, self-gravitating disks can develop 
GI even when $Q>1$. This is because viscosity compromises rotational
support against self-gravity for long-wavelength disturbances
\citep{lynden-bell74,willerding92,gammie96}. A similar effect 
occurs in dusty fluids where the required frictional forces are 
provided by dust-gas drag forces \citep{youdin11,takahashi14}.

Now, gaseous accretion disks are often modeled with a Navier-Stokes viscosity
to mimic hydrodynamic or magneto-hydrodynamic turbulence
\citep{shakura73}. This involves writing the
kinematic viscosity $\nu$ as  
\begin{align}\label{alpha_param}
  \nu  = \alpha \frac{c_s^2}{\Omega}, %dimensional argument 
\end{align}
where $\alpha$ is the dimensionless viscosity
coefficient. This
parameterization can be modified to include more complex dependencies on the
fluid variables, but Eq. \ref{alpha_param} is the general form.

How well turbulence can be modeled as an effective, Navier-Stokes
viscosity is a separate issue \citep{balbus99}. However, it is 
reasonable to assume that turbulence frustates rotational 
support. (This is, after all, the physical cause for turbulent mass
accretion.)

Again, viscosity is absent in the Toomre $Q$ parameter. The above
studies account for viscosity but consider isothermal disks. On the
other hand, proper inclusion of viscosity should also account for
viscous heating which, as discussed above, also affect stability. 
We thus extend the study of viscous GI to include an energy equation
with viscous heating and explicit cooling. This allows us to consider
GI in realistic PPDs.

@@

The basic equations are listed in \S\ref{basic}, along with
descriptions of the disk 
equilibra and the two cooling models 
considered --- parametrized `beta' cooling and radiative diffusion. 
The linear stability problem is defined in
\S\ref{linear}. We present results with beta cooling in
\S\ref{result_2d} and \S\ref{3ddisk} for 2D and 3D disks, respectively. In
\S\ref{2dppd} we apply linear theory to PPDs with realistic cooling
models, including radiative diffusion in 3D. We find regions in a
PPD that cannot sustain steady gravito-turbulence to coincide with
that expected for disk fragmentation, based on previous empirical
criteria. We summarize our results in \S\ref{summary}, including a
discussion of how our model may aid the physical understanding of  
fragmentation in realistic PPDs. 
% as the inability to sustain gravito-turbulence
%because the associated stress removes rotational support

>>>>

\begin{abstract}
Low mass, self-gravitating accretion disks can remain in a  
quasi-steady, `gravito-turbulent' state provided that thermal 
time-scales are sufficiently long, or equivalently for sufficiently
weak turbulent stresses. Otherwise, numerical simulations show that
disk fragmentation ensues. 
We quantify the interpretation of this result as the
compromise of gravitational stability by non-ideal physics: 
\begin{inparaenum}[1)] 
\item cooling, which reduces pressure support; and
\item viscosity, which reduces rotational support. 
\end{inparaenum}
To this end, we analyse the gravitational stability of accretion disks
including these effects explicitly.   
We derive a general dispersion relation for linear waves in a 
self-gravitating, viscous, non-adiabatic disk with internal and
external heating, and explicit cooling. We 
consider parameterized cooling functions in two and three dimensions,
as well as full radiative diffusion in 3D. 
We apply linear theory to 
global models of protoplanetary disks (PPDs) subject to gravito-turbulence, modeled as an effective
viscosity, and cooling via dust opacity. We find the sole effect of cooling is insufficient to
destabilize a realistic PPD. However, inclusion of viscosity renders
the disk beyond $\sim 60$AU dynamically unstable on radial 
lengthscales a few times the local disk thickness. This is coincident 
with the empirical condition for disk fragmentation based on a
maximum sustainable stress. We 
thus suggest that disk fragmentation can be physically attributed, 
at least partially, to the removal of rotational stabilization against
self-gravity by turbulence.   
\end{abstract}
%alternative abstract

\section{Introduction}\label{intro} 
The gravitational instability (GI) of rotating disks   
plays a fundamental role in many astrophysical problems. In the context of 
protoplanetary disks (PPDs), GI has two applications. It can 
provide gravitational torques to transport angular momentum outwards
and thus enable mass accretion inwards \citep{armitage10,turner14}.   
GI may also lead to disk fragmentation, which has been invoked to explain planet formation at
large radii \citep{helled14}. Any realistic study of these phenomena require direct
numerical simulations. Nevertheless, physical insight can be  
obtained through analytical understanding of GI.  %physical insight

@@

\subsection{Viscosity removes rotational support}
Viscous disks can develop GI even when $Q>1$. This is because
viscosity compromises rotational support against self-gravity for
long-wavelength disturbances
\citep{lynden-bell74,willerding92,gammie96}. A similar effect  
occurs in dusty fluids where the required frictional forces are 
provided by dust-gas drag \citep{ward00, takahashi14}. In fact, this
is a mechanism for planetesimational formation %frag of dust
\citep{youdin11}. Then it is not unreasonable to expect an analogous
fragmentation of gaseous disks due to viscosity.

For gaseous accretion disks, a Navier-Stokes viscosity is often used to 
to mimic hydrodynamic or magneto-hydrodynamic turbulence
\citep{shakura73}. This involves writing the
kinematic viscosity $\nu$ as  
\begin{align}\label{alpha_param}
  \nu  = \alpha \frac{c_s^2}{\Omega}, %dimensional argument 
\end{align}
where $\alpha$ is the dimensionless viscosity
coefficient. This
parameterization can be modified to include more complex dependencies on the
fluid variables, but Eq. \ref{alpha_param} is the general form.

How well turbulence can be modeled as an effective viscosity is a
separate issue \citep{balbus99}. However, it is  
reasonable to assume that turbulence frustates rotational 
support. (This is, after all, the physical cause for turbulent mass
accretion.) In this work, we use viscosity to represent this physical
effect.

Viscosity is absent in the standard Toomre parameter. The above
studies account for viscosity but consider isothermal disks. On the
other hand, proper inclusion of viscosity should also account for
viscous heating which, as discussed above, also affect stability. 
We thus extend the study of viscous GI to include an energy equation
with viscous heating and explicit cooling. This allows us to consider
GI in realistic PPDs.

@@

The basic equations are listed in \S\ref{basic}, along with
descriptions of the disk 
equilibra and the two cooling models 
considered --- parametrized `beta' cooling and radiative diffusion. 
The linear stability problem is defined in
\S\ref{linear}. We present results with beta cooling in
\S\ref{result_2d} and \S\ref{3ddisk} for 2D and 3D disks,
respectively. %key result: generalized dispersion relation 
In
\S\ref{2dppd} we apply linear theory to PPDs with realistic cooling
models, including radiative diffusion in 3D. We find regions in a
PPD that cannot sustain steady gravito-turbulence to coincide with
that expected for disk fragmentation, based on previous empirical
criteria. We summarize our results in \S\ref{summary}, including a
discussion of how our model may aid the physical understanding of  
fragmentation in realistic PPDs. 
% as the inability to sustain gravito-turbulence
%because the associated stress removes rotational support
